MG7: Configurable and scalable 16S metagenomics data analysis -- new methods optimized for massive cloud computing

### 2015-09-18 docs/paper/out/paper.tex
make wording changes in AWS M&M, review #35 -- paper
correct – to - --paper

During the past decade, metagenomics data analysis is growing
exponentially. Some of the reasons behind this are the increasing
throughput of massively parallel sequencing technologies (with the
derived decrease in sequencing costs), and the wide impact of
metagenomics studies \citep{oulas2015metagenomics}, especially in human
health (diagnostics, treatments, drug response or prevention)
\citep{bikel2015combining}. We should also mention what could be called
the microbiome explosion: all kind of microbiomes (gut, mouth, skin,
urinary tract, airway, milk, bladder) are now routinely sequenced in
different conditions of health and disease, or after different
treatments. The impact of Metagenomics is also being felt in
environmental sciences \citep{ufarte2015discovery}, crop sciences, the
agrifood sector \citep{coughlan2015biotechnological} and biotechnology
in general \citep[\citet{kodzius2015marine}]{cowan2015metagenomics}.
These new possibilities for exploring the diversity of micro--organisms
in the most varied environments are opening new research areas, and
drastically changing the existing ones.

@@

Genome researchers have raised the alarm over big data in the past
\citep{hayden2015genome}, but even a more serious challenge might be
faced with the metagenomics boom. If we compare metagenomics data with
other genomics data used in clinical genotyping we find a differential
feature: the key role of time. Thus, for example, in some longitudinal
studies, serial sampling from the same patient
\citep{faust2015metagenomics} along several weeks (or years) is being
used for the follow up of some intestinal pathologies, for studying the
evolution of the gut microbiome after antibiotic treatment, or for colon
cancer early detection
\citep[\citet{garrett2015cancer}]{zeller2014potential}. This need of
sampling across time adds more complexity to metagenomics data storage
and demands adapted algorithms to detect state variations across time as
well as idiosyncratic commonalities of the microbiome of each individual
\citep{franzosa2015identifying}. In addition to the intra-individual
sampling-time dependence, metagenomic clinical test results vary
depending on the specific region of extraction of the clinical specimen.
This local variability adds complexity to the analysis since different
localizations (different tissues, different anatomical regions, healthy
or tumour tissues) are required to have a sufficiently complete
landscape of the human microbiome. Moreover, re--analysis of old samples
using new tools and better reference databases might be also demanded
from time to time.

@@

Code compilation prior to launching any analysis assures that no AWS
resources are launched if the analysis is not well--defined, avoiding
expenses not leading to any analysis. Besides compile--time checks,
runtime checks are made before launch to ensure existence of input data
and availability of resources.

@@

MG7 uses various Amazon Web Services:

@@

These services are used through a Scala wrapper of the official
\href{https://aws.amazon.com/sdk-for-java/}{AWS Java SDK
\texttt{1.9.25}},
\href{https://github.com/ohnosequences/aws-scala-tools/releases/tag/v0.13.2}{ohnosequences/aws-scala-tools
\texttt{0.13.2}}.

>>>>

During the past decade, metagenomics data analysis is growing
exponentially. Some of the reasons behind this are the increasing
throughput of massively parallel sequencing technologies (with the
derived decrease in sequencing costs), and the wide impact of
metagenomics studies \citep{oulas2015metagenomics}, especially in human
health (diagnostics, treatments, drug response or prevention)
\citep{bikel2015combining}. We should also mention what could be called
the microbiome explosion: all kind of microbiomes (gut, mouth, skin,
urinary tract, airway, milk, bladder) are now routinely sequenced in
different conditions of health and disease, or after different
treatments. The impact of Metagenomics is also being felt in
environmental sciences \citep{ufarte2015discovery}, crop sciences, the
agrifood sector \citep{coughlan2015biotechnological} and biotechnology
in general \citep[\citet{kodzius2015marine}]{cowan2015metagenomics}.
These new possibilities for exploring the diversity of micro-organisms
in the most varied environments are opening new research areas, and
drastically changing the existing ones.

@@

Genome researchers have raised the alarm over big data in the past
\citep{hayden2015genome}, but even a more serious challenge might be
faced with the metagenomics boom. If we compare metagenomics data with
other genomics data used in clinical genotyping we find a differential
feature: the key role of time. Thus, for example, in some longitudinal
studies, serial sampling from the same patient
\citep{faust2015metagenomics} along several weeks (or years) is being
used for the follow up of some intestinal pathologies, for studying the
evolution of the gut microbiome after antibiotic treatment, or for colon
cancer early detection
\citep[\citet{garrett2015cancer}]{zeller2014potential}. This need of
sampling across time adds more complexity to metagenomics data storage
and demands adapted algorithms to detect state variations across time as
well as idiosyncratic commonalities of the microbiome of each individual
\citep{franzosa2015identifying}. In addition to the intra-individual
sampling-time dependence, metagenomic clinical test results vary
depending on the specific region of extraction of the clinical specimen.
This local variability adds complexity to the analysis since different
localizations (different tissues, different anatomical regions, healthy
or tumour tissues) are required to have a sufficiently complete
landscape of the human microbiome. Moreover, re-analysis of old samples
using new tools and better reference databases might be also demanded
from time to time.

@@

Code compilation prior to launching any analysis assures that no AWS
resources are launched if the analysis is not well-defined, avoiding
expenses not leading to any analysis. Besides compile-time checks,
runtime checks are made before launch to ensure existence of input data
and availability of resources.

@@

MG7 uses the following Amazon Web Services:

@@

These services are used through a Scala wrapper of the official
\href{https://aws.amazon.com/sdk-for-java/}{AWS Java SDK
\texttt{1.9.25}}:
\href{https://github.com/ohnosequences/aws-scala-tools/releases/tag/v0.13.2}{ohnosequences/aws-scala-tools
\texttt{0.13.2}}.
